\documentclass[11pt,a4paper]{article}

\usepackage[a4paper,margin=2.2cm]{geometry}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[table]{xcolor}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{array}
\usepackage{enumitem}
\usepackage{parskip}
\usepackage{hyperref}
\usepackage{titlesec}

\definecolor{accent}{HTML}{0F5C73}
\definecolor{soft}{HTML}{EEF6F8}
\definecolor{line}{HTML}{D3E5EB}
\definecolor{textgray}{HTML}{4E5B61}

\hypersetup{
  colorlinks=true,
  linkcolor=accent,
  urlcolor=accent
}

\titleformat{\section}{\Large\bfseries\color{accent}}{}{0em}{}
\titleformat{\subsection}{\large\bfseries\color{accent}}{}{0em}{}

\setlist[itemize]{leftmargin=1.4em, itemsep=3pt, topsep=4pt}
\renewcommand{\arraystretch}{1.2}

\newcommand{\note}[1]{%
  \begin{center}
  \fcolorbox{line}{soft}{%
    \parbox{0.95\linewidth}{#1}
  }
  \end{center}
}

\begin{document}

{\huge\bfseries Project Ideas\par}
\vspace{0.25em}
{\large\color{textgray}Shortlist focused on realistic projects with good chances of strong results\par}

\vspace{1em}

\section{Quick Ranking}

\rowcolors{2}{soft}{white}
\begin{tabularx}{\textwidth}{@{}>{\bfseries}c>{\raggedright\arraybackslash}p{4.1cm}>{\raggedright\arraybackslash}p{2.6cm}>{\raggedright\arraybackslash}p{2.0cm}X@{}}
\toprule
\textbf{\#} & \textbf{Idea} & \textbf{Dataset} & \textbf{Effort} & \textbf{Why it stands out} \\
\midrule
1 & Plant disease classification + Grad-CAM & PlantVillage & Low & Best balance between strong metrics, simple setup, and enough substance for the final report. \\
2 & Steel surface defect classification & NEU-CLS & Very low & Probably the safest option if we want something compact, quick to iterate on, and still meaningful. \\
3 & Traffic sign classification & GTSRB & Low & Very clean benchmark with good labels, strong baselines, and room for robustness experiments. \\
4 & Satellite land-use classification & EuroSAT & Low & Easy to train, visually interesting, and a bit more distinctive than a generic image dataset. \\
5 & Pneumonia classification from chest X-rays & Kermany Chest X-Ray & Low & Good demo value and clear metrics, but the ethics and limitations discussion needs more care. \\
\bottomrule
\end{tabularx}

\section{1. Plant Disease Classification + Grad-CAM}

It is a very safe project technically, but it still gives enough room to discuss model choice, explainability, class imbalance, and dataset bias.

\begin{itemize}
\item \textbf{Task:} classify plant leaf images into disease categories.
\item \textbf{Dataset:} PlantVillage, around 54k images and 38 classes.
\item \textbf{Model idea:} start with a simple CNN baseline, then fine-tune something light like ResNet18, MobileNetV3, or EfficientNet-B0.
\item \textbf{Evaluation:} accuracy, macro-F1, confusion matrix, per-class recall, and a few Grad-CAM visualizations.
\item \textbf{Why it is manageable:} public dataset, straightforward preprocessing, and transfer learning should already give strong results.
\item \textbf{Main risk:} the dataset is relatively clean and controlled, so real-world generalization may be weaker than the test results suggest.
\end{itemize}

\section{2. Steel Surface Defect Classification}

It is small, practical, and probably the fastest project to get working well.

\begin{itemize}
\item \textbf{Task:} classify steel surface images into defect categories.
\item \textbf{Dataset:} NEU-CLS, 1,800 grayscale images across 6 classes.
\item \textbf{Model idea:} simple CNN baseline, then ResNet18 or MobileNetV3 after converting grayscale to 3 channels.
\item \textbf{Evaluation:} accuracy, macro-F1, per-class recall, and ideally cross-validation because the dataset is small.
\item \textbf{Why it is manageable:} fast training, small dataset, clear labels, and an industrial application that still looks serious.
\item \textbf{Main risk:} because the dataset is small and visually consistent, the model may overfit to that specific setting.
\end{itemize}

\section{3. Traffic Sign Classification}

It is a classic benchmark, which makes the experimental setup very easy to justify.

\begin{itemize}
\item \textbf{Task:} classify cropped traffic sign images.
\item \textbf{Dataset:} GTSRB, more than 50k images and 40+ classes.
\item \textbf{Model idea:} custom CNN as baseline, then ResNet18 or EfficientNet-B0.
\item \textbf{Evaluation:} accuracy, macro-F1, confusion matrix, and maybe one robustness experiment with blur or brightness changes.
\item \textbf{Why it is manageable:} great dataset quality, lots of benchmark history, and no complicated pipeline.
\item \textbf{Main risk:} it is a bit standard, so the project becomes more interesting if we add robustness or efficiency analysis.
\end{itemize}

\section{4. Satellite Land-Use Classification}

It still feels simple, but the application domain is a bit more distinctive.

\begin{itemize}
\item \textbf{Task:} classify land-use categories from satellite image patches.
\item \textbf{Dataset:} EuroSAT, 27k labeled samples in 10 classes.
\item \textbf{Model idea:} ResNet18 or MobileNetV3, with EfficientNet-B0 as an optional extra comparison.
\item \textbf{Evaluation:} accuracy, macro-F1, confusion matrix, and error analysis for visually similar classes.
\item \textbf{Why it is manageable:} public dataset, low compute requirements, and easy transfer learning setup.
\item \textbf{Main risk:} it can become too straightforward, so the report should include a bit of analysis beyond raw accuracy.
\end{itemize}

\section{5. Pneumonia Classification from Chest X-Rays}

It has strong presentation value and very clear evaluation metrics, even with a simple binary classifier.

\begin{itemize}
\item \textbf{Task:} classify chest X-rays as normal or pneumonia.
\item \textbf{Dataset:} Kermany Chest X-Ray subset, around 5.8k labeled images.
\item \textbf{Model idea:} simple CNN baseline, then ResNet18 or DenseNet121, plus optional Grad-CAM.
\item \textbf{Evaluation:} accuracy, precision, recall, F1, AUROC, and special attention to false negatives.
\item \textbf{Why it is manageable:} binary classification keeps the pipeline simple and the metrics easy to interpret.
\item \textbf{Main risk:} this is the one that requires the most careful wording around bias, shortcuts, and real-world medical use.
\end{itemize}

\section{Best Picks Depending on What We Want}

\begin{itemize}
\item \textbf{Safest overall choice:} PlantVillage.
\item \textbf{Smallest implementation burden:} NEU-CLS.
\item \textbf{Best balance between simple and a bit more distinctive:} EuroSAT.
\item \textbf{Best benchmark-style project:} GTSRB.
\item \textbf{Best for a strong demo narrative:} Chest X-rays.
\end{itemize}

\section{Useful Links}

\begin{itemize}
\item PlantVillage: \url{https://www.tensorflow.org/datasets/catalog/plant_village}
\item NEU-CLS: \url{https://doi.org/10.6084/m9.figshare.28903550}
\item GTSRB: \url{https://benchmark.ini.rub.de/gtsrb_dataset.html}
\item EuroSAT: \url{https://github.com/phelber/EuroSAT}
\item Chest X-Ray dataset: \url{https://data.mendeley.com/datasets/rscbjbr9sj/3}
\end{itemize}

\end{document}
